---
title: "Predicting Individual Purchasing Behavior"
subtitle: "Full Distribution vs. Hurdle Models for Sales Strategy"
format: 
  revealjs:
    theme: default
    slide-number: true
    preview-links: auto
  pptx:
    # Remove the '#' on the next line and point it to a corporate PPTX file to use its template
    # reference-doc: my_corporate_template.pptx
    slide-level: 2
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Load necessary libraries
library(tidyverse)
library(scales)

# Set global ggplot text sizes (in mm) to match the PowerPoint target font size
update_geom_defaults("text", list(size = 5.5))

# Define the custom color palette
my_colors <- c(
    "Churn (Zero)" = "#D30F4B", # Raspberry
    "Decline"      = "#443247", # Dark Purple
    "Stable"       = "#00314E", # Dark Ocean Blue
    "Growth"       = "#56D500", # Deep Chartreuse
    "Capri"        = "#01BEFF" # Capri (Baseline)
)

# ---------------------------------------------------------
# 1. Data Generation (Individual Customer Prediction Scenario)
# ---------------------------------------------------------
set.seed(42)
n_simulations <- 10000
prob_zero <- 0.35 # ~35% probability of No Purchase

# Generate hurdle (0 or 1)
bought_anything <- rbinom(n_simulations, 1, 1 - prob_zero)

# Generate positive outcomes (normal around 80)
positive_amounts <- rnorm(n_simulations, mean = 80, sd = 15)

# Combine into a single predictive dataset
df <- tibble(
    bought = bought_anything == 1,
    amount = ifelse(bought, positive_amounts, 0)
)

# ---------------------------------------------------------
# 2. Key Metrics & Density Calculations
# ---------------------------------------------------------
p_zero <- mean(df$amount == 0)
p_pos <- 1 - p_zero

exp_total <- mean(df$amount) # E[Y]
exp_positive <- mean(df$amount[df$amount > 0]) # E[Y|Y>0]

# Spike dimensions (Area = p_zero)
spike_width <- 4
spike_height <- p_zero / spike_width

# Positive density (Area = p_pos)
pos_data <- df |>
    filter(amount > 0) |>
    pull(amount)
dens <- density(pos_data, from = min(pos_data), to = max(pos_data))

# Base density dataframe
dens_df <- tibble(
    amount = dens$x,
    density = dens$y * p_pos
)

# Add classification categories directly to the density data
last_year_val <- 80
dens_df <- dens_df |>
    mutate(
        Category = case_when(
            amount < (last_year_val * 0.95) ~ "Decline",
            amount >= (last_year_val * 0.95) & amount <= (last_year_val * 1.05) ~ "Stable",
            amount > (last_year_val * 1.05) ~ "Growth"
        ),
        Category = factor(Category, levels = c("Decline", "Stable", "Growth"))
    )

# Calculate exact proportions for text labels
prop_churn <- p_zero
prop_decline <- mean(df$amount > 0 & df$amount < (last_year_val * 0.95))
prop_stable <- mean(df$amount >= (last_year_val * 0.95) & df$amount <= (last_year_val * 1.05))
prop_growth <- mean(df$amount > (last_year_val * 1.05))
max_dens <- max(dens_df$density)

```

## The Anatomy of the Data

:::: {.columns}

::: {.column width="50%"}
**Full Prediction Distribution**

* Blends purchase likelihood and volume.
* Obscures the reality of how this client buys.

```{r}
#| fig-width: 5
#| fig-height: 3.5
df |>
    ggplot(aes(x = amount)) +
    geom_histogram(binwidth = 5, fill = my_colors["Stable"], color = "white") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    theme_minimal(base_size = 18) +
    theme(axis.line.x = element_line(color = "gray30", linewidth = 0.8)) +
    labs(x = "Predicted Order Volume", y = "Frequency")

```

:::

::: {.column width="50%"}
**Separated Positive Distribution**

* Isolates the "If they buy, how much?" question.
* Reveals a normal, predictable operational volume.

```{r}
#| fig-width: 5
#| fig-height: 3.5
df |>
    filter(amount > 0) |>
    ggplot(aes(x = amount)) +
    geom_histogram(binwidth = 5, fill = my_colors["Capri"], color = "white") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    theme_minimal(base_size = 18) +
    theme(axis.line.x = element_line(color = "gray30", linewidth = 0.8)) +
    labs(x = "Predicted Order Volume", y = "Frequency")

```

:::

::::

::: {.notes}
When we run our prediction model for an individual customer, the output isn't a single number. We are actually predicting two distinct behaviors: First, what is the probability this specific client will place an order at all? Second, if they do convert, what is their normal order volume? If we just look at historical averages, the reality of how this individual client buys is completely obscured. The hurdle model allows us to isolate those two behaviors so we can address them separately.
:::

## The Expectation Trap

* The blended expected value ($E[Y] \approx `r round(exp_total, 1)`$) falls in a gap.
* This customer will likely purchase around **`r round(exp_positive, 1)`** units, or they buy nothing.

```{r}
#| fig-width: 9
#| fig-height: 4
ggplot() +
    # The Zero Spike
    geom_col(aes(x = 0, y = spike_height), width = spike_width, fill = my_colors["Stable"], alpha = 0.85) +

    # The Positive Density Ribbon
    geom_ribbon(data = dens_df, aes(x = amount, ymin = 0, ymax = density), fill = my_colors["Capri"], alpha = 0.6) +
    geom_line(data = dens_df, aes(x = amount, y = density), color = my_colors["Stable"], linewidth = 1) +

    # Expectation Lines
    geom_vline(xintercept = exp_total, color = my_colors["Churn (Zero)"], linetype = "dashed", linewidth = 1.2) +
    geom_vline(xintercept = exp_positive, color = my_colors["Stable"], linewidth = 1.2) +

    # Annotations (Expectations)
    annotate("text",
        x = exp_total - 3, y = max_dens * 1.15,
        label = paste("Full Expectation\n~", round(exp_total, 1)),
        color = my_colors["Churn (Zero)"], fontface = "bold", hjust = 1, vjust = 0
    ) +
    annotate("text",
        x = exp_positive + 3, y = max_dens * 1.15,
        label = paste("Positive Expectation\n~", round(exp_positive, 1)),
        color = my_colors["Stable"], fontface = "bold", hjust = 0, vjust = 0
    ) +

    # Annotation (Individual Probability of No Purchase)
    annotate("text",
        x = spike_width + 2, y = spike_height * 0.85,
        label = paste0("Probability of\nNo Purchase\n(~", round(p_zero * 100), "%)"),
        color = "gray30", fontface = "italic", hjust = 0
    ) +

    # Force flush to the bottom, leave 15% padding at the top for labels
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +

    # Clean Styling & X-Axis Line
    theme_minimal(base_size = 18) +
    theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.line.x = element_line(color = "gray30", linewidth = 0.8)
    ) +
    labs(x = "Predicted Order Volume", y = "")

```

::: {.notes}
Here is the danger of handing a sales rep a single 'expected value' to guide their account plan. For this specific customer, the model indicates a 35% probability of no purchase—that is the solid spike on the left. However, if they do buy, their normal operational volume is around 80 units. If we mathematically blend those two facts, the system tells the rep to expect an order of ~52 units. The problem? This customer will never buy 52 units. Giving the rep the blended expectation sets them up to pitch the wrong volume to a client who might actually just need a basic retention push.
:::

## The Classification Shift

By mapping the account health categories directly onto the separated distributions (assuming a baseline of 80 units), we can clearly see our primary risk is strictly conversion (churn).

```{r}
#| fig-width: 9
#| fig-height: 4
ggplot() +
    # The Zero Spike (CHURN)
    geom_col(aes(x = 0, y = spike_height, fill = "Churn (Zero)"), width = spike_width, alpha = 0.95) +

    # The Positive Density Ribbon (Colored by Growth/Stable/Decline)
    geom_ribbon(data = dens_df, aes(x = amount, ymin = 0, ymax = density, fill = Category), alpha = 0.85) +

    # A thin line to define the curve's upper edge
    geom_line(data = dens_df, aes(x = amount, y = density), color = "white", linewidth = 0.5) +

    # Text Overlays for Proportions (added white boxes for better contrast)
    annotate("label", x = 0, y = spike_height * 0.5, label = percent(prop_churn, 1), color = "gray15", fill = "white", fontface = "bold", size = 4.5, label.size = 0, label.padding = unit(0.3, "lines")) +
    annotate("label", x = last_year_val * 0.8, y = max_dens * 0.2, label = percent(prop_decline, 1), color = "gray15", fill = "white", fontface = "bold", size = 4.5, label.size = 0, label.padding = unit(0.3, "lines")) +
    annotate("label", x = last_year_val, y = max_dens * 0.4, label = percent(prop_stable, 1), color = "gray15", fill = "white", fontface = "bold", size = 4.5, label.size = 0, label.padding = unit(0.3, "lines")) +
    annotate("label", x = last_year_val * 1.25, y = max_dens * 0.2, label = percent(prop_growth, 1), color = "gray15", fill = "white", fontface = "bold", size = 4.5, label.size = 0, label.padding = unit(0.3, "lines")) +

    # Apply the custom color palette
    scale_fill_manual(
        name = "Account Health\n(Relative to Y_t-1 = 80)",
        values = my_colors
    ) +

    # Force the bottom of the plot to sit exactly at y = 0
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +

    # Clean Styling & X-Axis Line
    theme_minimal(base_size = 18) +
    theme(
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(),
        legend.position = "right",
        axis.line.x = element_line(color = "gray30", linewidth = 0.8)
    ) +
    labs(x = "Predicted Order Volume", y = "")

```

::: {.notes}
Let's say this customer bought 80 units last year. How do we flag their account health in the CRM? If we use the blended expectation of 52, our dashboard flags them as 'Declining'. A rep seeing a 'Declining' tag might panic and offer unnecessary volume discounts. But look at the hurdle model's separated reality: if we get them to buy, their volume is highly likely to remain 'Stable' or even hit 'Growth'. The real risk isn't that their operational needs are shrinking; the risk is the 35% probability they churn entirely. The rep's goal shouldn't be negotiating order size—it must be purely focused on securing the renewal.
:::


## The Illusion of Decline

:::: {.columns}

::: {.column width="100%"}

```{r}
#| fig-width: 10
#| fig-height: 6

# Plot A: The Naive Business Rule (The Illusion)
# Here, the zero-spike is treated as just a very severe "Decline"
p_illusion <- ggplot() +
    geom_col(aes(x = 0, y = spike_height, fill = "Decline"), width = spike_width, alpha = 0.95) +
    geom_ribbon(data = dens_df, aes(x = amount, ymin = 0, ymax = density, fill = Category), alpha = 0.85) +
    geom_line(data = dens_df, aes(x = amount, y = density), color = "white", linewidth = 0.5) +
    geom_vline(xintercept = last_year_val, color = "black", linetype = "dashed") +
    annotate("text",
        x = last_year_val + 5, y = max_dens * 0.8,
        label = "Previous Year\n(80 Units)", hjust = 0, fontface = "italic"
    ) +
    scale_fill_manual(values = my_colors) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_minimal(base_size = 12) +
    theme(
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(),
        axis.line.x = element_line(color = "gray30", linewidth = 0.8),
        legend.position = "none" # Hide legend to keep focus on the colors
    ) +
    labs(
        title = "1. Naive Business Rules: Treating Non-Buyers as 'Decline'",
        subtitle = "Because 0 is less than 80, the system paints churned accounts and shrinking active accounts with the same brush.",
        x = "", y = ""
    )

# Plot B: The Hurdle Reality (Isolating Churn)
# Here, the exact same zero-spike is isolated as "Churn"
p_reality <- ggplot() +
    geom_col(aes(x = 0, y = spike_height, fill = "Churn (Zero)"), width = spike_width, alpha = 0.95) +
    geom_ribbon(data = dens_df, aes(x = amount, ymin = 0, ymax = density, fill = Category), alpha = 0.85) +
    geom_line(data = dens_df, aes(x = amount, y = density), color = "white", linewidth = 0.5) +
    geom_vline(xintercept = last_year_val, color = "black", linetype = "dashed") +
    scale_fill_manual(values = my_colors) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_minimal(base_size = 12) +
    theme(
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(), panel.grid.minor.y = element_blank(),
        axis.line.x = element_line(color = "gray30", linewidth = 0.8),
        legend.position = "bottom",
        legend.title = element_blank()
    ) +
    labs(
        title = "2. Hurdle Model Reality: Isolating Churn Risk",
        subtitle = "By separating the hurdle, we reveal the true health of our active buyers.",
        x = "Predicted Order Volume", y = ""
    )

# Use patchwork to stack them cleanly
library(patchwork)
p_illusion / p_reality
```

:::

::::

::: {.notes}
Look at the top chart. If we feed the full, un-separated distribution into our CRM and apply our health categories, the massive weight of the zero-buyers drags the entire curve to the left. The system flags almost the entire portfolio as 'Declining' (the dark purple). If I am a sales leader looking at this, I am firing my account managers because it looks like every single customer is slowly abandoning us.

Now look at the bottom chart—the exact same underlying data, but processed through the hurdle model. The truth is revealed. We do not have a systemic 'Decline' problem. Our active buyers (the blue and green areas) are healthy, stable, and growing. Our only actual problem is the raspberry wall on the left: Churn.

This changes everything. We shouldn't be giving broad volume discounts to save 'declining' accounts; we should be launching a targeted win-back campaign for the 35% who are walking out the door entirely.
:::


## Portfolio Segmentation: The Two-Metric Advantage

:::: {.columns}

::: {.column width="60%"}

```{r}
#| fig-width: 6.5
#| fig-height: 5

# Generate synthetic territory data (e.g., 150 accounts a rep manages)
set.seed(123)
territory_df <- tibble(
    account_id = 1:150,
    prob_buy = runif(150, 0.2, 0.95), # Probability of purchase (1 - P_zero)
    exp_vol = rnorm(150, mean = 80, sd = 25) # Conditional expected volume
) |>
    # Define the quadrants based on median or business-logic splits
    mutate(
        Action = case_when(
            prob_buy < 0.6 & exp_vol >= 80 ~ "Flight Risk VIPs\n(Intervene)",
            prob_buy >= 0.6 & exp_vol >= 80 ~ "Core Pipeline\n(Protect & Upsell)",
            prob_buy < 0.6 & exp_vol < 80 ~ "Low Priority\n(Automate)",
            prob_buy >= 0.6 & exp_vol < 80 ~ "Steady Builders\n(Expand Volume)"
        ),
        Action = factor(Action, levels = c("Flight Risk VIPs\n(Intervene)", "Core Pipeline\n(Protect & Upsell)", "Steady Builders\n(Expand Volume)", "Low Priority\n(Automate)"))
    )

# Define quadrant colors using your palette
quad_colors <- c(
    "Flight Risk VIPs\n(Intervene)" = my_colors[["Churn (Zero)"]], # Raspberry
    "Core Pipeline\n(Protect & Upsell)" = my_colors[["Growth"]], # Chartreuse
    "Steady Builders\n(Expand Volume)" = my_colors[["Capri"]], # Capri
    "Low Priority\n(Automate)" = my_colors[["Decline"]] # Dark Purple
)

territory_df |>
    ggplot(aes(x = prob_buy, y = exp_vol, color = Action)) +

    # Background quadrant shading (subtle)
    annotate("rect", xmin = -Inf, xmax = 0.6, ymin = 80, ymax = Inf, fill = my_colors[["Churn (Zero)"]], alpha = 0.05) +
    annotate("rect", xmin = 0.6, xmax = Inf, ymin = 80, ymax = Inf, fill = my_colors[["Growth"]], alpha = 0.05) +
    annotate("rect", xmin = -Inf, xmax = 0.6, ymin = -Inf, ymax = 80, fill = my_colors[["Decline"]], alpha = 0.05) +
    annotate("rect", xmin = 0.6, xmax = Inf, ymin = -Inf, ymax = 80, fill = my_colors[["Capri"]], alpha = 0.05) +

    # Quadrant divider lines
    geom_hline(yintercept = 80, linetype = "dashed", color = "gray50") +
    geom_vline(xintercept = 0.6, linetype = "dashed", color = "gray50") +

    # The scatter points
    geom_point(size = 3, alpha = 0.8) +

    # Apply colors
    scale_color_manual(values = quad_colors) +

    # Formatting
    scale_x_continuous(labels = percent) +
    theme_minimal(base_size = 18) +
    theme(
        legend.position = "bottom",
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        legend.margin = margin(t = -10)
    ) +
    labs(
        x = "Likelihood to Purchase",
        y = "Expected Conditional Volume"
    )
```

:::

::: {.column width="40%"}

<br>
<br>

**The Dimensions of Execution**

* A single blended average only allows reps to sort accounts from large to small.
* Hurdle models provide two specific dimensions, unlocking precise account segmentation.
* Immediately identifies **Flight Risks** vs **Steady Builders**, ensuring time is spent where it yields highest return.

:::

::::

::: {.notes}
By unblending the expected value, we don't just get better math; we get better execution. If we only give a sales rep a single, blended metric, they treat a stable 50-unit buyer exactly the same as a 100-unit buyer who has a 50% chance of churning.

But when we map the two hurdle metrics into a matrix, it becomes a strategic goldmine. The rep knows exactly where to spend their time. The Raspberry accounts in the top left are our 'Flight Risk VIPs'—they require immediate, high-touch phone calls. Meanwhile, 'Low Priority' accounts in the bottom left can be pushed to an automated marketing drip. This ensures a rep's time is spent where it matters most: protecting the core pipeline and expanding the steady builders.
:::

## Strategic Alignment

| Feature | Full Distribution $E[Y]$ | Hurdle Separated $P(Y=0)$ & $E[Y|Y>0]$ |
| :--- | :--- | :--- |
| **Mathematical Reality** | Blends conversion risk with operational volume. | Isolates conversion probability from order magnitude. |
| **Prediction Output** | A smoothed, blended volume that misrepresents the client. | Two distinct metrics: Likelihood to buy, and conditional volume. |
| **Best Used For** | **Macro-level forecasting:** Aggregating total projected revenue across thousands of accounts. | **Micro-level execution:** Individual account planning, targeted marketing, and setting rep quotas. |
| **The Risk** | Under-pitching actual buyers, or misinterpreting stable clients as shrinking. | Requires reps to manage two metrics instead of one. |

::: {.notes}
Ultimately, these two mathematical approaches serve different masters. The Full Distribution is for the aggregate business—it is what our supply chain and finance teams use, because across thousands of customers, the individual churn probabilities smooth out into predictable revenue. But for Sales execution, we must use the Separated Hurdle approach. We need to arm our reps with two distinct metrics per account: the Conversion Risk to drive retention tactics, and the Conditional Volume to drive their quota and upsell targets.
:::
